---
title: "Getting Started with OmicsBatchR"
author: "Panomiq Bioinformatics Team"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with OmicsBatchR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 10,
  fig.height = 6
)
```

# Introduction

**OmicsBatchR** provides a single, unified interface to run and benchmark
multiple batch correction methods for single-cell and bulk RNA-seq data.

Batch effects are a common problem in omics experiments — samples processed
in different labs, on different days, or with different protocols often show
technical variation that masks the real biology. Correcting for batch effects
is essential before clustering, differential expression, or any downstream analysis.

OmicsBatchR supports these methods:

| Method | Type | Best for |
|--------|------|----------|
| Harmony | R | Large datasets, fast |
| Seurat CCA | R | Multi-dataset integration |
| ComBat-seq | R | Count-level bulk RNA-seq |
| FastMNN | R | Single-cell, conservative |
| Scanorama | Python | Panoramic multi-batch |
| BBKNN | Python | Very large datasets |
| scVI | Python | Deep learning, complex batches |

# Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("panomiq/OmicsBatchR")

# For Python-based methods, also install:
reticulate::conda_create("r-reticulate")
reticulate::conda_install("r-reticulate",
  packages = c("scvi-tools", "scanorama", "bbknn", "scanpy", "anndata"),
  pip = TRUE)
```

# Quick start

## Load the package

```{r load}
library(OmicsBatchR)
library(Seurat)
```

## Prepare your Seurat object

OmicsBatchR requires a Seurat object with a batch variable in the metadata.

```{r prepare, eval=FALSE}
# Your Seurat object should have a 'batch' column
head(your_seurat@meta.data)
#   orig.ident nCount_RNA nFeature_RNA batch cell_type
#   Cell1      2500       800          Batch1 T_cell
#   Cell2      3100       950          Batch2 B_cell
```

## Run all batch correction methods

```{r run_all, eval=FALSE}
results <- run_batch_correction(
  seurat_obj = your_seurat,
  batch_key  = "batch",
  methods    = "all",      # runs all 7 methods
  n_dims     = 30,
  verbose    = TRUE
)
```

## Run specific methods only

```{r run_specific, eval=FALSE}
results <- run_batch_correction(
  seurat_obj = your_seurat,
  batch_key  = "batch",
  methods    = c("harmony", "fastmnn", "seurat_cca"),
  n_dims     = 30
)
```

# Visualise results

## UMAP comparison grid

```{r plot_umap, eval=FALSE}
# Colour by batch — check how well batches are mixed
plot_batch_comparison(
  results,
  color_by = "batch"
)

# Colour by cell type — check biological signal is preserved
plot_batch_comparison(
  results,
  color_by = c("batch", "cell_type"),
  ncol     = 4
)

# Save to PDF
plot_batch_comparison(
  results,
  color_by  = "batch",
  save_path = "batch_comparison_umap.pdf",
  width     = 18,
  height    = 12
)
```

# Benchmark methods

## Compute LISI and kBET scores

```{r benchmark, eval=FALSE}
scores <- benchmark_methods(
  results,
  batch_key    = "batch",
  celltype_key = "cell_type"   # optional: for cLISI
)

print(scores)
#   method    iLISI  cLISI  kBET_acceptance  overall_score
#   harmony   1.85   1.12   0.78             0.91
#   fastmnn   1.72   1.09   0.71             0.84
#   scvi      1.91   1.08   0.82             0.93
#   ...
```

**Metric interpretation:**

- **iLISI** — Integration LISI. Measures batch mixing. *Higher is better.*
  Range: 1 (no mixing) to n_batches (perfect mixing).
- **cLISI** — Cell-type LISI. Measures cell-type separation. *Lower is better.*
  Range: 1 (perfect separation) to n_celltypes.
- **kBET acceptance rate** — Local neighbourhood batch composition test.
  *Higher is better.* Range: 0–1.
- **Overall score** — Weighted composite (iLISI 40% + cLISI 30% + kBET 30%).
  *Higher is better.*

## Plot benchmark scores

```{r plot_scores, eval=FALSE}
plot_benchmark_scores(scores)
```

# Auto-select the best method

```{r select_best, eval=FALSE}
best <- select_best_method(
  results,
  batch_key    = "batch",
  celltype_key = "cell_type"
)

cat("Best method:", best$best_method)
# Best method: scvi

# The corrected Seurat object is ready for downstream analysis
best_obj <- best$seurat_obj
```

# Downstream analysis on best-corrected object

```{r downstream, eval=FALSE}
best_obj <- best$seurat_obj

# Find neighbours using corrected embedding
best_obj <- FindNeighbors(best_obj, reduction = "scvi", dims = 1:30)
best_obj <- FindClusters(best_obj, resolution = 0.5)

# Differential expression
markers <- FindAllMarkers(best_obj, only.pos = TRUE, min.pct = 0.25)
```

# Python environment setup

For Scanorama, BBKNN, and scVI (Python-based methods):

```{r python_setup, eval=FALSE}
# One-time setup
reticulate::conda_create("r-reticulate", python_version = "3.9")
reticulate::conda_install("r-reticulate",
  packages = c("scvi-tools", "scanorama", "bbknn", "scanpy", "anndata"),
  pip = TRUE
)

# Pass the environment name to run_batch_correction
results <- run_batch_correction(
  seurat_obj = your_seurat,
  batch_key  = "batch",
  methods    = c("scvi", "scanorama", "bbknn"),
  python_env = "r-reticulate"
)
```

# Session info

```{r session}
sessionInfo()
```
